Optimal (Multiway) Spatial Joins

### 2024-06-03 msj.tex
yufei

In terms of techniques, our primary contribution is the revelation of a new property on the problem's mathematical structure. Fix any $k \ge 3$ and an {\em arbitrary} algorithm $\A$ for the $(k-1)$-SJ problem. Define function $F_{k-1}(n, \out)$ to return the worst-case running time of $\A$ on any instance of the $(k-1)$-SJ problem having input size at most $n$ and output size at most $\out$. Our core technical contribution is to establish:

\begin{theorem} \label{thm:main-recur}
    Equipped with the algorithm $\A$ as described above, the $k$-SJ problem with $k \ge 3$ can be solved in time
    \myeqn{
        O(k^3) \cdot \big( F_{k-1}(n, \out) + n \log n + k \cdot \out \big)
        %\nn
        \label{eqn:main:reccurrence}
    }
    where $n$ (resp., $\out$) is the input (resp., output) size of the problem. Furthermore, if $\A$ is comparison-based, so is the $k$-SJ algorithm.
\end{theorem}

The theorem implies a recursive nature of $k$-SJ. Indeed, we will see that an $k$-SJ instance with input size $n$ and output size $\out$ can be converted to $O(k^3)$ instances of the $(k-1)$-SJ problem --- all having input size at most $n$ and output size at most $\out$ --- plus an additional processing cost of $O(n \log n + \out)$. For 2-SJ, we can set $\A$  to the ``folklore algorithm'' mentioned in Section~\ref{sec:intro:prev}, which ensures $F_2(n, \out) = O(n \log n + \out)$. Combining this with \eqref{eqn:main:reccurrence} gives a recurrence that relates the time complexity of $k$-SJ to that of $(k-1)$-SJ. Solving the recurrence yields:

\begin{theorem} \label{thm:main-alg}
    For $k \ge 3$, we can settle $k$-SJ with a comparison-based algorithm in $O( c^k \cdot (k!)^3 \cdot (n \log n + k \cdot \out))$ time, where $c > 1$ is a positive constant.
    %and $O(k \cdot n + \out)$ space,
\end{theorem}

When $k = O(1)$, the time complexity becomes $O(n \log n$ $+ \out)$, as promised; the space consumption of our algorithm is $O(n + \out)$. Now that Theorem~\ref{thm:main-alg} offers a satisfactory $k$-SJ result for $k = O(1)$ in 2D space, it is natural to wonder whether the constraint on dimensionality 2 is necessary. Interestingly, the answer is ``yes'' as far as $k \ge 3$ is concerned, subject to the absence of breakthroughs on a classical problem in graph theory. Specifically, if the 3D version of the 3-SJ problem (which we will formally define in Appendix~\ref{app:lb-cond}) could be solved in $O((n + \out) \cdot \polylog n)$ time, we would obtain an algorithm that detects the presence of a triangle (i.e., 3-clique) in a graph of $m$ edges in $O(m \polylog m)$ time, which would make a truly remarkable breakthrough because the state of the art needs $O(m^{1.41})$ time \cite{ayz97}. This reduction can be inferred from an argument in \cite{kcko22} used to prove a more generic result. We simplify the argument for 3D 3-SJ and present the full reduction in Appendix~\ref{app:lb-cond}.
\section{Preliminaries in Geometry} \label{sec:bricks}

>>>>

Our primary technical contribution is the revelation of a new property on the problem's mathematical structure. Fix any $k \ge 3$ and an {\em arbitrary} algorithm $\A$ for the $(k-1)$-SJ problem. Define function $F_{k-1}(n, \out)$ to return the worst-case running time of $\A$ on any instance of the $(k-1)$-SJ problem having input size at most $n$ and output size at most $\out$. We will establish:

\begin{theorem} \label{thm:main-recur}
    Equipped with the algorithm $\A$ as described above, the $k$-SJ problem with $k \ge 3$ can be solved in time
    \myeqn{
        O(k^3) \cdot \big( F_{k-1}(n, \out) + n \log n + k \cdot \out \big)
        %\nn
        \label{eqn:main:reccurrence}
    }
    where $n$ (resp., $\out$) is the input (resp., output) size of the problem. Furthermore, if $\A$ is comparison-based, the obtained $k$-SJ algorithm is also comparison-based.
\end{theorem}

The theorem implies a recursive nature of $k$-SJ. Indeed, we will see that an $k$-SJ instance with input size $n$ and output size $\out$ can be converted to $O(k^3)$ instances of the $(k-1)$-SJ problem --- all having input size at most $n$ and output size at most $\out$ --- plus an additional processing cost of $O(n \log n + k \cdot \out)$. For 2-SJ, we can set $\A$  to the ``folklore algorithm'' mentioned in Section~\ref{sec:intro:prev}, which ensures $F_2(n, \out) = O(n \log n + \out)$. Combining this with \eqref{eqn:main:reccurrence} gives a recurrence that relates the time complexity of $k$-SJ to that of $(k-1)$-SJ. Solving the recurrence yields:

\begin{theorem} \label{thm:main-alg}
    For $k \ge 3$, we can settle $k$-SJ with a comparison-based algorithm in
    \myeqn{
        O( c^k \cdot (k!)^3 \cdot (n \log n + k \cdot \out))
        \nn
    }
    time, where $c > 1$ is a positive constant.
    %and $O(k \cdot n + \out)$ space,
\end{theorem}

When $k = O(1)$, the time complexity becomes $O(n \log n$ $+ \out)$, as promised; the space consumption of our algorithm is $O(n + \out)$. Now that Theorem~\ref{thm:main-alg} offers a satisfactory $k$-SJ result for $k = O(1)$ in 2D space, it is natural to wonder whether the constraint on dimensionality 2 is necessary. Interestingly, the answer is ``yes'' as far as $k \ge 3$ is concerned, subject to the absence of breakthroughs on a classical problem in graph theory. Specifically, if the 3D version of the 3-SJ problem (which we will formally define in Appendix~\ref{app:lb-cond}) could be solved in $O((n + \out) \cdot \polylog n)$ time, we would be able to detect the presence of a triangle (i.e., 3-clique) in a graph of $m$ edges in $O(m \polylog m)$ time, which would make a remarkable breakthrough because the state of the art needs $O(m^{1.41})$ time \cite{ayz97}. This reduction can be inferred from an argument in \cite{kcko22} used to prove a more generic result. We simplify the argument for 3D 3-SJ and present the full reduction in Appendix~\ref{app:lb-cond}.

\section{Preliminaries in Geometry} \label{sec:bricks}
